Software Quality Assurance in INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures

### 2019-09-25 paper-jgc.tex
reliability term usage

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe through dedicated software development projects for
e--Infrastructure creation, operation and management.
This work presents the most significant software quality breakthroughs obtained in
one of such projects, {\sl INDIGO--DataCloud}, the main challenges and barriers 
confronted throughout the lifespan of the project, and how they were partially or 
totally overcome.
The knowledge base established throughout the last 15 years 
of diverse software development initiatives in Europe for sustaining distributed research 
e-Infrastructures, supported by the advances in the area of software engineering, definitely
%constituted a significant breakthrough in the reliability of the software delivered in the 
contributed to improve the reliability of the software delivered, and consequently, the 
operational stability of the European e--Infrastructures. {\sl INDIGO--DataCloud} project is 
a good evidence of such insights, where, unlike the preceding trend found in past projects,
the enforcement of Software Quality Assurance practices has been present since the very early 
stages of the software lifecycle.
\keywords{Software Reliability \and Quality Assurance \and Software Metrics \and Software Testing
Techniques \and DevOps}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

Throughout the last 15 years, the European Commission (EC) has continuously invested on software development initiatives to provide
research e--Infrastructures with the capability of supporting unified access to large--scale computing and intense 
data analysis facilities. The developed tools have enabled the federation of geographically distributed resource 
providers, with the ultimate aim of supporting the growing needs of the scientific comumunities. According to their
requirements, those research e--Infrastructures have been evolving, exploiting diverse distributed computing 
technologies namely Grid, in the first place, and, eventually, Cloud.

The urging needs of providing stable e--Infrastructures, which were based on novel distributed computing technologies, 
initially drove the efforts spent in software development towards the provision of new functionalities, reducing the 
adoption of software engineering methodologies to a marginal player. The released software was less tested, resulting in 
a high rate of rollbacks and patches applied to production systems, when compared to the subsequent projects. 
Thereby, the stability and reliability of the e--Infrastructures were often compromised \cite{lingrand-egee}, leading 
to outages in the participant resource centers. Back in the days when Grid middleware was actively developed, 
researchers reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, 
eventually developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.

In an effort to improve the operational experience of research communities, it soon became apparent the need for a better 
balance between the addition of new features and the reliability of the released software. Consequently, incoming European 
funding programmes and open calls progressivelly stressed the fact of devoting a more substantial part of the software 
lifecycle to the improvement of the quality and sustainability of the products being delivered. At this point,
evolving software engineering practices were considered and gradually adopted in order to define and implement the quality
procedures, the organization of cross--functional teams and the deployment of pilot testbeds.

@@

Following the DevOps principles, the SQA criteria requires the verification of the source code on each change in the 
production version, defining the essential requirements to be processed. Such scenario is hardly accomplished without
the aid of automation, and as a consequence -- as described in Section~\ref{software-automation} -- the testing and 
validation solution adopted by {\sl INDIGO--DataCloud} relied on the great advancements that continuous integration 
(CI) tools have undergone in the last years. Accordingly, both the validation and delivery of {\sl INDIGO--DataCloud} 
software was tackled by automation means.

The fundamental shift brought by the {\sl INDIGO--DataCloud} SQA procedures was 
to act at the early stages of the software development lifecycle. Thus, instead of validating the software product once 
it was packaged within a pre-release stage, the source code itself is verified for each minor change. This strategy 
allows to improve the sustainability of the software, by checking the compliance with code style standards, and its 
reliability, through the execution of functional and integration tests, ensuring the total absence of regressions. The results
presented in this paper show a divergence between the high rate of bug fixes resolved at development time and the 
significantly lower number of bugs detected by the end users. Consequently, bugs are more effortessly fixed since they 
are early captured and end users interface with more reliable and stable software.

The paper is organized as follows. %Section~\ref{sec:b} summarizes the essential context of the research problem.
Section~\ref{sec:ev} details the evolution of research on software reliability in the selected EC--projects. 
Section~\ref{sec:ntsr} provides information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project. Conclusions are drawn in Section~\ref{sec:con}.

@@

One of the fundamental pillars of the {\sl INDIGO--DataCloud}'s SQA process has been established
by the outcomes resulted from similar EC--funded e--Infrastructure--related projects over
the last 15 years. In particular, in this section we are primarily interested in hightlighting the 
collaborative
software development efforts within the category of projects that improved the accessibility and
exploitation of distributed computing technologies in the European research landscape. As such, 
we will present the most representative software engineering practices adopted by those projects
to understand how they ultimately inspired the SQA process definition and implementation in 
{\sl INDIGO--DataCloud}, further discussed in Section~\ref{sec:ntsr}.

@@

Project monitoring and reporting mechanisms allowed to assess the risks, efforts and documentation
produced. Development guidelines \cite{datagridguide} were produced, covering the various phases, including
1) packaging, 2) test and validation, and 3) style and naming convention. The specification of quality indicators
\cite{datagridindicators} facilitated the detection of deviations in the collected metric data. Figure
\ref{fig:efficiency} shows the efficiency performance indicator, measured in production environments as the
number of jobs successfully completed divided by the total number of jobs submitted. As it can be observed,
there is a positive trend of efficiency over the months, with the unique exception of the LCG--1 prototype.
\begin{figure}[h]
\centering
%\includegraphics[width=0.8\textwidth, height=50mm]{images/datagridEff.png}
\includegraphics[width=1\textwidth]{images/datagridEff.png}
\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are provided for five experiments involved in the project.}
\label{fig:efficiency}
\end{figure}

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee,cordis:egee2,cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities~\cite{egee}. The software to sustain the increasing
requirements coming from the diverse scientific communities needed developing a
rich set of new services, while maintaining a sustainable infrastructure for
Grid computing. This e-Infrastructure was eventually used by more than 15 thousand researchers
and deployed in over 250 institutions.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{images/fig2.png}
\caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud}
software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases. Products that
were exclusively released during {\sl INDIGO--2} do not show data for the first period of
{\sl INDIGO--1}.}
\label{fig:fig_unittest}
\end{figure*}

@@

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the definition and periodic execution
of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development
activities. \textit{Unit testing} completes the source code testing evaluation by focusing on the code's
internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage
for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software
component between the two major releases, denoting an incremental trend over the course of the project:
code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
the project's first (INDIGO--1) and second (INDIGO--2) releases. Nevertheless, Fig.~\ref{fig:fig_unittest}
shows a decrease in the unit testing coverage values for 18\% of products. As unit tests cover low-level
code elements, they are best considered while the new code is being written \cite{unit-test-frameworks}.
We have observed that this decrease was aligned with high-demanding periods of software development, as
seen in the previous months of the INDIGO-2 major release. Regardless of the potentially higher defect
density that this fact brought along, the overall performance kept progressing as the rest of the software
stack substantially improved their own unit testing coverage statistics. By INDIGO--2, 53\% of the software
components (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage
recommendation (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.
The individual records followed also a growing trend (only 6 out of 32 components had lower values in
INDIGO--2). It is worth noting that Fig.~\ref{fig:fig_unittest} not only includes new software developments
implemented from scratch within the project. Software tools and libraries from
external open--source projects -- such as {\sl tosca-parser} or {\sl heat-translator} --, contributed upstream
by the project, and well--established products, involved in {\sl INDIGO--DataCloud}
but not contributing with the 100\% of their codebases -- such as {\sl dCache} --, are also considered in the
analysis. However the values given apply to the entire codebase, even for the latter case of products. These
products challenged the application of the SQA policies described in this section, by two means:
\begin{itemize}
    \item Previously developed code was not refactored since it was not \textit{owned} by the project.
    \item An agreement on prevailing SQA policies for such cases where quality practices were already in place.
\end{itemize}

@@

\subsection{Software quality validation and delivery automation}\label{software-automation}
\label{sec:devops}
Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
is hardly achieved without the aid of automation. The introduction of automation to validate the quality
requirements increases the overall reliability of the produced software: automated testing is more
time-efficient, leading to higher code coverages and increased defect detection, when compared with a manual
approach \cite{rafi2012}. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
total number of defects detected in the pre--production stages surpassed by a factor of 30
the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud}
project.
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth, height=50mm]{images/indigoBugs.png}
\caption{Number of software bugs -- documentation typos are excluded -- detected in development 
and production stages over the lifetime of the {\sl INDIGO-DataCloud} project. Production 
defects correspond to software bugs filed by users of the EGI e--Infrastructure throughout 
the next 2 years after the {\sl INDIGO--1} major release, in August 2016 (source: EGI's GGUS
helpdesk). The development defects stem only from the pre--release testing phase, filed 
exclusively as part of the SQA team work (source: GitHub Issues tracker). As a consequence, the 
development defects started to arise in the previous months prior to the {\sl INDIGO--1} release, 
while the major activity was concentrated in the months that preceded the second major release 
of {\sl INDIGO--2}, in April 2017. As it can be observed, the offset between the advent of defects
in development and production corresponds to nearly half a year, with the first production defect
reported 3 months after the {\sl INDIGO--1} release.}

@@

As any new change, involving a new feature or fix, passes through the automated
SQA machinery, the software is always ready to be released. As a result, the often compromised harmonization
between the continuous release of new features -- pushed by the development teams -- and the maintenance of
stable production systems -- demanded by users and resource providers -- can be guaranteed. The DevOps culture
theorizes and provides practical solutions that aim at unifying both software development (dev) and software
operations (ops) teams, by emphasizing the SQA techniques to avoid infrastructure disruption whenever new
developments are deployed into production systems.
The incremental adoption of DevOps approaches throughout the project's lifetime was a key player towards
software reliability. Starting with a Continuous Integration (CI) scenario at the first stages of the project, the software delivered
was eventually validated and distributed using Continuous Delivery (CD) pipelines.

@@

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the {\sl INDIGO--DataCloud}
solutions and use cases validation. Within these testbeds, the applications are deployed
and executed
with the last stable version of the software in environments as similar as
possible to production ones. A map of the Pilot Preview
infrastructure is depicted in Fig.~\ref{fig:fig_pilotpreview}. It shows the
resource providers and the components or services they deployed and supported.

The released software is tested in production environments through the
Staged--Rollout process, before its inclusion in the EGI's Cloud Middleware Distribution (CMD).
Selected resource providers are requested to install
the most updated stable versions of the software, and to give access to them to their users. The
Staged--Rollout process is key to detect and mitigate issues that could only
appear in production environments.

@@

All along the project we have observed that source code review is the cornerstone to
improve the effectiveness of the unit and functional -- when applicable -- test cases. 
{\sl INDIGO--DataCloud} source code repositories are protected against direct pushes 
so that the code review stage always takes place. Code reviewers analyse the content
of each change in the source code and the suitability -- according to the SQA criteria -- of the
associated tests, being the last checkpoint before the change is added into the
production branch. Hence the vital importance of providing the tests alongside the
code for every given change.
% ?? figure with code reviews or sth about PRs, external participation
%Functional quality, on the other hand, was assessed within the {\sl INDIGO--DataCloud} 
%software lifecycle through the per--change execution of the functional tests provided 
%by the developers and the ultimate stage of integration testing in the aforementioned 
%preview testbed. As we have seen, the latter testing strategy was entirely performed 
%in a manual way, and appeared as a helpful tool to spot defects after exposing the 
%deployed software to interactions with external components. Additionally, the functional 
%tests, were primarily useful to check that the software met the design specifications, 
%and most importantly, that previously working functionalities are preserved when new 
%features are added. As mentioned, these type of testing is meant to be continuously 
%executed in an automated manner using the CI infrastructure. 
However, by the time of
{\sl INDIGO--1} only 30\% of the total of products provided automated functional tests,
reaching almost 60\% in {\sl INDIGO--2}. Although test reports were asked in those cases
where automated functional tests were not provided, the fulfillment of 
their fundamental functional requirements was tougher to track, resulting in a non--viable
solution for assessing minor changes, especially when analysing the regressions. The major issues observed that hindered developers from the 
implementation of automated functional tests were releated to the unavailability
-- at the time -- of libraries and frameworks for tackling specific tests -- such as 
testing graphical interfaces --, or in some cases, the unawareness about their existence.
We infer from this experience that automated functional tests should be a goal in modern
software development as they complement the structural quality checks by adding the
validation of the software's identified functionalities, key in the operational impact 
and user acceptance. 
% figure with % of products with functional tests
% ?? products not developed from scratch challenged the application of SQA policies??

@@

@@

@@

>>>>

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe through dedicated software development projects for
e--Infrastructure creation, operation and management.
This work presents the most significant software quality breakthroughs obtained in
one of such projects, {\sl INDIGO--DataCloud}, the main challenges and barriers 
confronted throughout the lifespan of the project, and how they were partially or 
totally overcome.
The knowledge base established throughout the last 15 years 
of diverse software development initiatives in Europe for sustaining distributed research 
e-Infrastructures, supported by the advances in the area of software engineering, definitely
%constituted a significant breakthrough in the reliability of the software delivered in the 
contributed to improve the quality and reliability of the software delivered, and consequently, the 
operational stability of the European e--Infrastructures. {\sl INDIGO--DataCloud} project is 
a good evidence of such insights, where, unlike the preceding trend found in past projects,
the enforcement of Software Quality Assurance practices has been present since the very early 
stages of the software lifecycle.
\keywords{Software Reliability \and Quality Assurance \and Software Metrics \and Software Testing
Techniques \and DevOps}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

Throughout the last 15 years, the European Commission (EC) has continuously invested on software development initiatives to provide
research e--Infrastructures with the capability of supporting unified access to large--scale computing and intense 
data analysis facilities. The developed tools have enabled the federation of geographically distributed resource 
providers, with the ultimate aim of supporting the growing needs of the scientific communities. According to their
requirements, those research e--Infrastructures have been evolving, exploiting diverse distributed computing 
technologies namely Grid, in the first place, and, eventually, Cloud.

The urging needs of providing stable e--Infrastructures, which were based on novel distributed computing technologies, 
initially drove the efforts spent in software development towards the provision of new functionalities, reducing the 
adoption of software engineering methodologies to a marginal player. The released software was less tested, resulting in 
a high rate of rollbacks and patches applied to production systems, when compared to the subsequent projects. 
Thereby, the stability of the e--Infrastructures were often compromised \cite{lingrand-egee}, leading 
to outages in the participant resource centers. Back in the days when Grid middleware was actively developed, 
researchers reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, 
eventually developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.

In an effort to improve the operational experience of research communities, it soon became apparent the need for a better 
balance between the addition of new features and the quality of the released software. Consequently, incoming European 
funding programmes and open calls progressively stressed the fact of devoting a more substantial part of the software 
lifecycle to the improvement of the quality and sustainability of the products being delivered. At this point,
evolving software engineering practices were considered and gradually adopted in order to define and implement the quality
procedures, the organization of cross--functional teams and the deployment of pilot testbeds.

@@

Following the DevOps principles, the SQA criteria encourages a thorough verification of the source code, inspecting
and testing each minor change meant to be included in the production version. Considering that 
{\sl INDIGO--DataCloud} solution comprises more than 30 software products and 200 source code repositories, such 
scenario is hardly accomplished without the aid of automation. Consequently, the SQA process builds on the great 
advancements that the continuous integration (CI) tools have undergone over the last years, as described in 
Section~\ref{software-automation}. Hence, acting at the early stages of the software development lifecycle brought
a fundamental shift when compared with the reviewed {\sl INDIGO--DataCloud}'s preceding projects that dealt with
the task of developing software solutions for Grid and Cloud--based e--Infrastructures.

This strategy 
allows to improve the overall quality of the software, both in terms of sustainability, by checking the compliance with code style standards, and 
reliability, through the execution of functional and integration tests, ensuring the total absence of regressions. The results
presented in this paper show a divergence between the high rate of bug fixes resolved at development time and the 
significantly lower number of bugs detected by the end users. Consequently, bugs are more effortlessly fixed since they 
are early captured and end users interface with more reliable and stable software.

The paper is organized as follows. %Section~\ref{sec:b} summarizes the essential context of the research problem.
Section~\ref{sec:ev} details the evolution of SQA processes in the preceding EC--funded projects that developed 
software solutions for Grid and Cloud--based e--Infrastructures. 
Section~\ref{sec:ntsr} thoroughly describes the new trends in software engineering conducted during the course of the recently finished
{\sl INDIGO--DataCloud} project. Conclusions are drawn in Section~\ref{sec:con}.

@@

One of the fundamental pillars of the {\sl INDIGO--DataCloud}'s SQA process has been established
by the outcomes resulted from similar EC--funded e--Infrastructure--related projects over
the last 15 years. In particular, in this section we are primarily interested in emphasizing the 
collaborative
software development efforts within the category of projects that improved the accessibility and
exploitation of distributed computing technologies in the European research landscape. As such, 
we will present the most representative software engineering practices adopted by those projects
to understand how they ultimately inspired the SQA process definition and implementation in 
{\sl INDIGO--DataCloud}, further discussed in Section~\ref{sec:ntsr}.

@@

Project monitoring and reporting mechanisms allowed to assess the risks, efforts and documentation
produced. Development guidelines \cite{datagridguide} were produced, covering the various phases, including
1) packaging, 2) test and validation, and 3) style and naming convention. A set of quality indicators
\cite{datagridindicators} were defined to measure the efficiency of the system from the user perspective. 
Due to the developmental nature of DataGrid project, they solely consider the \textit{crude efficiency} 
metric as the relevant indicator to track the quality and performance of the core software, also known as
\emph{middleware}
in Grid terminology. This metric considered the success rate of all the jobs submitted by users using the 
production testbed, so that all the failures of the middleware were included in the statistic \cite{avellino2004datagrid}. However, it
also accounted for the failures in the remaining pieces in the job submission chain, such as the user application 
software or the targeted datacenter deployment. Even so, as it can be observed in Fig.~\ref{fig:efficiency}, 
the crude efficiency showed a positive trend over the months that followed, with the unique exception of the 
LCG--1 middleware distribution.
\begin{figure}[h]
\centering
%\includegraphics[width=0.8\textwidth, height=50mm]{images/datagridEff.png}
\includegraphics[width=1\textwidth]{images/datagridEff.png}
\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are 
provided for the different DataGrid's EDG middleware distributions released throughout the project. Note that the figure
includes metrics of the same release (EDGv1.4) for three Large Hadron Collider (LHC) experiments.}
\label{fig:efficiency}
\end{figure}

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee,cordis:egee2,cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities~\cite{egee}. The software to sustain the increasing
requirements coming from the diverse scientific communities needed developing a
rich set of new services, while maintaining a sustainable infrastructure for
Grid computing. This e-Infrastructure was eventually used by more than 15 thousand researchers
and deployed in over 250 institutions \cite{ferrari2011resources}.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{images/fig2.png}
\caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud}
software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases. Products that
were exclusively released during {\sl INDIGO--2} do not show data for the first period of
{\sl INDIGO--1}. Source code for all the {\sl INDIGO--DataCloud} components can be found
in the GitHub's \emph{indigo--dc} organization \cite{foot9}}
\label{fig:fig_unittest}
\end{figure*}

@@

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the definition and periodic execution
of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development
activities. \textit{Unit testing} completes the source code testing evaluation by focusing on the code's
internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage
for unit testing. In Figure.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software
component between the two major releases, denoting an incremental trend over the course of the project:
code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
the project's first (INDIGO--1) and second (INDIGO--2) releases. Nevertheless, Fig.~\ref{fig:fig_unittest}
shows a decrease in the unit testing coverage values for 18\% of products. As unit tests cover low-level
code elements, they are best considered while the new code is being written \cite{unit-test-frameworks}.
We have observed that this decrease was aligned with high-demanding periods of software development, as
seen in the previous months of the INDIGO-2 major release. Regardless of the potentially higher defect
density that this fact brought along, the overall performance kept progressing as the rest of the software
stack substantially improved their own unit testing coverage statistics. By INDIGO--2, 53\% of the software
components (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage
recommendation (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.
The individual records followed also a growing trend (only 6 out of 32 components had lower values in
INDIGO--2). It is worth noting that Fig.~\ref{fig:fig_unittest} not only includes new software developments
implemented from scratch within the project. Software tools and libraries from
external open--source projects -- such as {\sl tosca-parser} or {\sl heat-translator} --, contributed upstream
by the project, and well--established products, involved in {\sl INDIGO--DataCloud}
but not contributing with the 100\% of their codebases -- such as {\sl dCache} --, are also considered in the
analysis. However the values given apply to the entire codebase, even for the latter case of products. These
products challenged the application of the SQA policies described in this section, by two means:
\begin{itemize}
    \item Previously developed code was not refactored since it was not \textit{owned} by the project.
    \item An agreement on prevailing SQA policies for such cases where quality practices were already in place.
\end{itemize}

@@

\subsection{Software quality validation and delivery automation}\label{software-automation}
\label{sec:devops}
Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
is hardly achieved without the aid of automation. The introduction of automation to validate the quality
requirements increases the overall reliability of the produced software: automated testing is more
time-efficient, leading to higher code coverages and increased defect detection, when compared with a manual
approach \cite{rafi2012}. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
total number of defects detected in the pre--production stage, i.e. throughout the integration testing phase, surpassed by a factor of 30
the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud}
project. An accurate estimation of the software bugs detected -- and fixed -- with the aid of 
the CI/CD infrastructure, previously conducted at the source code analysis stage, cannot be
given as those bugs are usually fixed on the fly without being tracked. Nevertheless, we can
safely state that they exceed by far the total amount of 200 bugs that the Fig.~\ref{fig:indigobugs}
shows.
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth, height=50mm]{images/indigoBugs.png}
\caption{Number of software bugs -- documentation typos are excluded -- detected in development 
and production stages over the lifetime of the {\sl INDIGO-DataCloud} project. Production 
defects correspond to software bugs filed by users of the EGI e--Infrastructure throughout 
the next 2 years after the {\sl INDIGO--1} major release, in August 2016 (source: EGI's GGUS
helpdesk). The development defects stem only from the pre--release testing phase, filed 
exclusively as part of the SQA team work (source: GitHub Issues tracker). As a consequence, the 
development defects started to arise in the previous months prior to the {\sl INDIGO--1} release, 
while the major activity was concentrated in the months that preceded the second major release 
of {\sl INDIGO--2}, in April 2017. As it can be observed, the offset between the advent of defects
in development and production corresponds to nearly half a year, with the first production defect
reported 3 months after the {\sl INDIGO--1} release.}

@@

As any new change, involving a new feature or fix, passes through the automated
SQA machinery, the software is always ready to be released. As a result, the often compromised harmonization
between the continuous release of new features -- pushed by the development teams -- and the maintenance of
stable production systems -- demanded by users and resource providers -- can be guaranteed. The DevOps culture
theorizes and provides practical solutions that aim at unifying both software development (dev) and software
operations (ops) teams, by emphasizing the SQA techniques to avoid infrastructure disruption whenever new
developments are deployed into production systems. Starting with a Continuous Integration (CI) scenario at the first stages of the project, the software delivered
was eventually validated and distributed using Continuous Delivery (CD) pipelines.

@@

Two pilot infrastructures, integration and preview, were at the disposal of developers and use cases, 
respectively, involved in the project. The aim of these testbeds is to test the level of integration 
between the core {\sl INDIGO--DataCloud}'s components and the user applications.

The integration testbed was primarily targeted and used by the software developers. 
There, unstable
software releases were tested by the maintainers themselves in order to validate the interactions
with coupled services. Computing and storage resources were occasionaly needed, meaning that no real
testbed -- with all
the {\sl INDIGO--DataCloud}'s components -- was needed.

Conversely, the preview testbed was actively maintained, where the last stable version of
the software was deployed by the SQA team. On top of that, the user applications were 
deployed in order to test the integration with the {\sl INDIGO--DataCloud} solution. The
preview testbed proved to be an effective tool to uncover bugs by means of testing the 
basic functionalities of the core components. The bugs tagged as \emph{Dev} in
Fig.~\ref{fig:indigobugs} were spotted this stage. A map of the Pilot Preview
infrastructure is depicted in Fig.~\ref{fig:fig_pilotpreview}. It shows the
resource providers and the services deployed. The number of resources
steadily increased as the project progressed, due mainly to the progressive increase of 
supported services and the occasional need for deploying more than one stable version of
the same software component in different providers.

Computing and storage resource consumption, required by the pilot testbeds and SQA services
(metrics, CI/CD infrastructure), was never an issue for the project's resource providers.
In the case of computing resources, the massive use of container--based virtualization 
alleviated the need of resources as it allowed CPU and memory over--subscription, when no
performance test was carried out. Most of the providers were already members of the EGI 
FedCloud, providing medium--sized Infrastructure--as--a--Service Cloud frameworks with 
enough computing and storage capacity to meet the deployment requirements of more than 30 
products that comprised the {\sl INDIGO--DataCloud} solution and the related SQA services.
Nevertheless, the pilot infrastructure coordination task distributed the deployments
according each provider's capacity and effort received within the project. Consequently,
Fig.~\ref{fig:fig_pilotpreview} shows an unbalanced distribution of services.

As the final validation step, the released software was tested in production environments
through the \emph{staged--rollout} process, before its inclusion in the EGI's Cloud 
Middleware Distribution (CMD). This process selected resource providers from the EGI FedCloud
infrastructure in order to install the last stable versions of the {\sl INDIGO--DataCloud}
software, giving user access to them. The staged--rollout process is key to detect and 
mitigate issues that could only appear in production environments.

@@

All along the project we have observed that source code review is the cornerstone to
improve the effectiveness of the unit and functional -- when applicable -- test cases. 
{\sl INDIGO--DataCloud} source code repositories are protected against direct pushes 
so that the code review stage always takes place. Code reviewers analyse the content
of each change in the source code and the suitability -- according to the SQA criteria -- of the
associated tests, being the last checkpoint before the change is added into the
production branch. Hence the vital importance of providing the tests alongside the
code for every given change.
% ?? figure with code reviews or sth about PRs, external participation
%Functional quality, on the other hand, was assessed within the {\sl INDIGO--DataCloud} 
%software lifecycle through the per--change execution of the functional tests provided 
%by the developers and the ultimate stage of integration testing in the aforementioned 
%preview testbed. As we have seen, the latter testing strategy was entirely performed 
%in a manual way, and appeared as a helpful tool to spot defects after exposing the 
%deployed software to interactions with external components. Additionally, the functional 
%tests, were primarily useful to check that the software met the design specifications, 
%and most importantly, that previously working functionalities are preserved when new 
%features are added. As mentioned, these type of testing is meant to be continuously 
%executed in an automated manner using the CI infrastructure. 
However, by the time of
{\sl INDIGO--1} only 30\% of the total of products provided automated functional tests,
reaching almost 60\% in {\sl INDIGO--2}. Although test reports were asked in those cases
where automated functional tests were not provided, the fulfillment of 
their fundamental functional requirements was tougher to track, resulting in a non--viable
solution for assessing minor changes, especially when analysing the regressions. The major issues observed that hindered developers from the 
implementation of automated functional tests were related to the unavailability
-- at the time -- of libraries and frameworks for tackling specific tests -- such as 
testing graphical interfaces --, or in some cases, the unawareness about their existence.
We infer from this experience that automated functional tests should be a goal in modern
software development as they complement the structural quality checks by adding the
validation of the software's identified functionalities, key in the operational impact 
and user acceptance. 
% figure with % of products with functional tests
% ?? products not developed from scratch challenged the application of SQA policies??

@@

\bibitem{avellino2004datagrid}
G. Avellino \emph{et al.},
``The DataGrid workload management system: Challenges and results,''
\emph{J. Grid Comp.},
vol. 2, no. 4, pp. 353--367, 2004.

@@

\bibitem{ferrari2011resources}
T. Ferrari \emph{et al.},
``Resources and services of the EGEE production infrastructure,''
\emph{J. Grid Comp.},
vol. 9, no. 2, pp. 119--133, 2011.

@@

\bibitem{foot9}
GitHub's indigo-dc organization, ``indigo-dc,''
2018.
[Online]. Available: \url{https://github.com/indigo-dc}
[Accessed 14 Feb. 2019].
